---
title: 'Alternative Splicing Analysis on SRA or User-Owned Data'
author: "Nuno Saraiva-Agostinho"
date: "`r Sys.Date()`"
bibliography: refs.bib
csl: bioinformatics.csl
output: 
    rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{SRA or user-owned RNA-seq data analysis}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

---

*psichomics* is an interactive R package for integrative analyses of alternative
splicing using data from [The Cancer Genome Atlas (TCGA)][tcga] (containing 
molecular data associated with 34 tumour types), the
[Genotype-Tissue Expression (GTEx)][gtex] project (containing data for multiple
normal human tissues), the [Sequence Read Archive][SRA] or owned by the user. 

The following tutorial goes through the steps required to load custom RNA-seq 
data:

1. Retrieve FASTQ files and sample-associated information from the 
[Sequence Read Archive (SRA)][sra] (optional if you already have your own FASTQ 
samples);
2. Map RNA-seq reads from the FASTQ files against a genome of reference using
[STAR][star], a splice-aware aligner;
3. Merge and prepare the output of such aligners to be correctly interpreted
by psichomics;
4. Load data into psichomics.

# Download data from SRA (optional)

[SRA][sra] is a repository of biological sequence data that stores data from 
many published articles. [SRA][sra] data may be useful to answer pressing 
biological questions using publicly available data.

> The latest versions of psichomics support 
**automatic downloading of SRA data** from [recount2][recount], a resource of
pre-processed data for thousands of SRA projects (including gene read counts, 
splice junction quantification and sample metadata). Check first if the project 
of your interest is not available through this resource, thus making it easier 
to analyse gene expression and alternative splicing for your samples of 
interest.

Data from SRA can be downloaded using the [fastq-dump][fastq-dump] command from
[sra-tools][sra-tools]. For instance, to retrieve samples from the 
[SRP126561][SRP126561] project, we could do the following:

```
fastq-dump --gzip --split-3 SRR6368612
fastq-dump --gzip --split-3 SRR6368613
fastq-dump --gzip --split-3 SRR6368614
fastq-dump --gzip --split-3 SRR6368615
fastq-dump --gzip --split-3 SRR6368616
fastq-dump --gzip --split-3 SRR6368617
```

Arguments used in the previous command:

- **`--gzip`**: compress FASTQ files in the GZIP format
- **`--split-3`**: allows to output one or two FASTQ files for single-end or
paired-end sequencing, respectively (a third FASTQ file may also be returned 
containing orphaned single-end reads obtained from paired-end sequencing data)

Sample-associated data is also available from the [Run Selector][SRP126561]
page. Click **RunInfo Table** to download the whole metadata table for all
samples.

# Align RNA-seq data to quantify splice junctions

The quantification of each alternative splicing event is based on the proportion
of junction reads that support the inclusion isoform, known as percent 
spliced-in or PSI [@wang2008].

To estimate this value for each splicing event, both alternative splicing
annotation and quantification of RNA-Seq reads aligning to splice junctions 
(junction quantification) are required. While alternative splicing annotation is
provided by the package, junction quantification will need to be prepared from
user-owned data by aligning the RNA-seq reads from FASTQ files to a genome of
reference. As junction reads are required to quantify alternative splicing, a
splice-aware aligner will be used. psichomics currently supports STAR output.

## STAR

Before aligning FASTQ samples against a genome of reference, a genome index 
needs to be prepared. 

Start by downloading a FASTA file of the whole genome and a GTF file with
annotated transcript. To run the following example, download the 
[human FASTA and GTF files (hg19 assembly)][hg19-data].

```
mkdir hg19_STAR
STAR --runThreadN 4 \
     --runMode genomeGenerate \
     --genomeDir hg19_STAR \
     --genomeFastaFiles /path/to/hg19.fa \
     --sjdbGTFfile /path/to/hg19.gtf
```

Arguments used in the previous command:

- **`--runThreadN 4`**: run STAR in parallel using 4 threads
- **`--runMode genomeGenerate`**: generate the genome index
- **`--genomeDir`**: directory of the genome index directory (output)
- **`--genomeFastaFiles`**: path to genome file (FASTA format)
- **`--sjdbGTFfile`**: path to transcript annotation (GTF format)

---

After the genome index is generated, align the FASTQ files using the following
commands to align read counts to both genes and splice junctions. These commands
will allow STAR to output.

```
STAR --runThreadN 4 \
     --genomeDir hg19_STAR \
     --quantMode GeneCounts \
     --readFilesCommand zcat \
     --outFileNamePrefix SRR6368612 \
     --readFilesIn SRR6368612_1.fastq.gz SRR6368612_2.fastq.gz
     
STAR --runThreadN 4 \
     --genomeDir hg19_STAR \
     --quantMode GeneCounts \
     --readFilesCommand zcat \
     --outFileNamePrefix SRR6368613 \
     --readFilesIn SRR6368613_1.fastq.gz SRR6368613_2.fastq.gz
     
STAR --runThreadN 4 \
     --genomeDir hg19_STAR \
     --quantMode GeneCounts \
     --readFilesCommand zcat \
     --outFileNamePrefix SRR6368614 \
     --readFilesIn SRR6368614_1.fastq.gz SRR6368614_2.fastq.gz
     
STAR --runThreadN 4 \
     --genomeDir hg19_STAR \
     --quantMode GeneCounts \
     --readFilesCommand zcat \
     --outFileNamePrefix SRR6368615 \
     --readFilesIn SRR6368615_1.fastq.gz SRR6368615_2.fastq.gz
     
STAR --runThreadN 4 \
     --genomeDir hg19_STAR \
     --quantMode GeneCounts \
     --readFilesCommand zcat \
     --outFileNamePrefix SRR6368616 \
     --readFilesIn SRR6368616_1.fastq.gz SRR6368616_2.fastq.gz
     
STAR --runThreadN 4 \
     --genomeDir hg19_STAR \
     --quantMode GeneCounts \
     --readFilesCommand zcat \
     --outFileNamePrefix SRR6368617 \
     --readFilesIn SRR6368617_1.fastq.gz SRR6368617_2.fastq.gz
```

Arguments used in the previous command:

- **`--runThreadN 4`**: run STAR in parallel using 4 threads
- **`--genomeDir`**: directory of the genome index directory (input)
- **`--quantMode GeneCounts`**: count reads aligning per gene
- **`--readFilesCommand`**: command to extract compressed FASTQ files
- **`--outFileNamePrefix`**: output prefix
- **`--readFilesIn`**: FASTQ files to align

# Prepare output for psichomics

To process the resulting data files, open an R console or RStudio and type:

```{r, eval=FALSE}
# Change working directory to where the STAR output is
setwd("/path/to/aligned/output/")

library(psichomics)
prepareGeneQuant(
    "SRR6368612ReadsPerGene.out.tab", "SRR6368613ReadsPerGene.out.tab",
    "SRR6368614ReadsPerGene.out.tab", "SRR6368615ReadsPerGene.out.tab",
    "SRR6368616ReadsPerGene.out.tab", "SRR6368617ReadsPerGene.out.tab")
prepareJunctionQuant("SRR6368612SJ.out.tab", "SRR6368613SJ.out.tab", 
                     "SRR6368614SJ.out.tab", "SRR6368615SJ.out.tab",
                     "SRR6368616SJ.out.tab", "SRR6368617SJ.out.tab")
prepareSRAmetadata("SraRunTable.txt")
```

## Load data in visual interface

Start psichomics with the following commands in an R console or RStudio:

```{r, eval=FALSE}
library(psichomics)
psichomics()
```

Then, click **Load user files**. Click the **Folder input** tab and select the
appropriate folder where the psichomics-prepared data is stored. Finally, click
**Load files** to scan the folder for data that may be loaded by psichomics.

## Load data in command-line interface (CLI)

```{r, eval=FALSE}
library(psichomics)
data          <- loadLocalFiles("/path/to/aligned/output/")
geneExpr      <- data[[1]]$`Gene expression`
junctionQuant <- data[[1]]$`Junction quantification`
sampleInfo    <- data[[1]]$`Sample metadata`
```

# Feedback

All feedback on the program, documentation and associated material (including
this tutorial) is welcome. Please send any comments and questions to:

> Nuno Saraiva-Agostinho (nunoagostinho@medicina.ulisboa.pt)
>
> [Disease Transcriptomics Lab, Instituto de Medicina Molecular (Portugal)][distrans]

# References

[tcga]: https://tcga-data.nci.nih.gov/docs/publications/tcga
[gtex]: https://www.gtexportal.org
[sra]: https://www.ncbi.nlm.nih.gov/sra
[star]: https://github.com/alexdobin/STAR
[distrans]: http://imm.medicina.ulisboa.pt/group/distrans/
[fastq-dump]: https://ncbi.github.io/sra-tools/fastq-dump.html
[sra-tools]: https://ncbi.github.io/sra-tools/
[hg19-data]: https://grch37.ensembl.org/info/data/ftp/
[SRP126561]: https://trace.ncbi.nlm.nih.gov/Traces/study/?acc=SRP126561
[recount]: https://jhubiostatistics.shinyapps.io/recount/