---
title: "Differential analysis"
output: html_document
---

# Splicing
**Adapted from Marie's and Lina's Tutorials**

To perform differential analysis on alternative splicing, we need the exon
inclusion levels ("PSI" tables). Let's load them.

```{r creating PSI table}
# Create PSI table from
INCLUSION_LEVELS_FULL.Hsa8 <- read.delim("~/Dropbox/FCUL/Tese de Mestrado/Supporting Materials/INCLUSION_LEVELS_FULL-Hsa8.tab")

psi <- INCLUSION_LEVELS_FULL.Hsa8[-c(1:5, seq(6, 22, by=2))]
rownames(psi) <- INCLUSION_LEVELS_FULL.Hsa8[, 2]

# Scaling PSIs from 0 to 1 insted of 0 to 100
psi <- psi/100
```

Exclude events that do not vary across samples.
```{r remove variance of 0 or NA}
var_events <- rowVar(psi, na.rm = TRUE)
psi <- psi[!(is.na(var_events) | var_events == 0), ]
```

Quantify the percentage of NAs per event: let's only consider events with up to
40% of NAs (more than 50% is really not acceptable).

```{r remove high percentage of NAs}
nasPerRow <- rowSums(is.na(psi))
rows <- ncol(psi)
na_events <- nasPerRow/rows

# Plot histogram of NAs per row
library(ggplot2)
ggplot(psi, aes(x=na_events)) + 
    geom_histogram(colour="black", fill="light blue") + 
    xlab("NA fraction") + ylab("Count") + ggtitle("NA content per Event")

psiFiltered <- psi[na_events < 0.4, ]
```

# Statistical tests

## Non-parametric tests
Assuming the data does not follow normal distribution, non-parametric tests are
required. Such kind of tests include:

- **Data dependency:** Spearman's rank correlation coefficient (quantitative
outcomes) and Fisher's exact test
- **Median tests:** Wilcoxon rank-sum test (two condition comparison; also known
as Mann-Whitney U Test) and Wilcoxon signed-rank test (non-parametric version of
the paired two condition comparison's *t*-test)
- **Variance tests:** Kruskal-Wallis rank sum test (multiple condition 
comparison) and Levene's test

# Plots
The two types of plots most used in differential analysis are boxplots and 
density plots. Boxplots are available in the package `Highcharter` by using the 
function `hc_add_series_boxplot` (currently only available in the development 
version at GitHub):
```{r}
library(dplyr)
library(purrr)
library(stringr)

hc_add_series_df <- function(hc, data, ...) {
    
    assertthat::assert_that(is.data.frame(data))
    
    hc <- hc %>%
        hc_add_series(data = list.parse3(data), ...)
    
    hc
}

#' Shorcut to create a boxplot
hc_add_series_boxplot <- function(hc, x, by = NULL, outliers = TRUE, ...) {
    if (is.null(by)) {
        by <- "value"
    } else {
        stopifnot(length(x) == length(by))
    }
    
    df <- data_frame(value = x, by = by) %>% 
        group_by(by) %>% 
        do(data = boxplot.stats(.$value))
    
    bxps <- map(df$data, "stats")
    
    hc <- hc %>%
        hc_xAxis(categories = df$by) %>% 
        hc_add_series(data = bxps, type = "boxplot", ...)
    
    if (outliers) {
        outs <- map2_df(seq(nrow(df)), df$data, function(x, y){
            if (length(y$out) > 0)
                d <- data_frame(x = x - 1, y = y$out)
            else
                d <- data_frame()
            d
        })
        
        if (nrow(outs) > 0) {
            hc <- hc %>%
                hc_add_series_df(
                    data = outs,
                    name = str_trim(paste(list(...)$name, "outliers")),
                    type = "scatter", #linkedTo = ":previous",
                    marker = list(...),
                    tooltip = list(
                        headerFormat = "<span>{point.key}</span><br/>",
                        # pointFormat = "Observation: {point.y}"
                        pointFormat = 
                            "<span style='color:{point.color}'></span> Outlier: <b>{point.y}</b><br/>"
                    ),
                    ...
                )
        }
        
        
    }
    return(hc)
}

library(highcharter)
data(diamonds, package = "ggplot2")

highchart() %>% 
    hc_legend(enabled = FALSE) %>%
    hc_add_series_boxplot(diamonds$x, diamonds$color,
                          name = "X", color = "#2980b9") 
```

Regarding density plots, there are no supporting functions. Therefore, I created
a function to easily create them:

```{r}
hc_add_series_density <- function (hc, x, area = FALSE, ...) {
    type <- ifelse(area, "area", "spline")
    data <- list.parse3(data.frame(cbind(x = x$x, y = x$y)))
    return(hc %>% hc_add_series(data = data, type = type))
}

d1 <- density(mtcars$mpg[1:32/2])
d2 <- density(mtcars$mpg[32/2:32])
d3 <- density(mtcars$mpg[5:(16+5)])

highchart() %>% 
    hc_add_series_density(d1, area = TRUE) %>% 
    hc_add_series_density(d2, area = TRUE) %>% 
    hc_add_series_density(d3, area = TRUE)
```

# References
- https://en.wikipedia.org/wiki/Student%27s_t-test
- [rSeqNP: A non-parametric approach for detecting differential ex-
pression and splicing from RNA-Seq data](http://bioinformatics.oxfordjournals.org/content/31/13/2222)